---
title: "Using Archive Data"
author: "Chase Mateusiak"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  Some notes on using data from the `lts` archive
  brentlabRnaSeqTools package version: `r packageVersion("brentlabRnaSeqTools")`
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Using Archive Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```

```{r setup}
library(brentlabRnaSeqTools)
```

# Using data from the lts archive

From 20220923 forward, I am going to carefully version both the archive and 
the `brentlabRnaSeqTools`. In the past, I haven't done that, and as the database 
has changed, so has the codebase used to parse and interpret it. As a result, 
the archived data may not always work with the most current version of this 
package. I can help you access the appropriate version/commit in the git history 
if you need it. Alternatively, since the data is clean and you're a data scientist, 
you should be able to parse the frames on your own outside of this package.

That said, for more recent archived versions, it should work. To use the archive, 
do something like the following:

```{r}
library(brentlabRnaSeqTools)
library(tidyverse)

# mount the cluster, or download the 20220208 kn99 db archive
# paths will need to be updated for your machine. These are examples
# of what it would look like if you mount.
archive_prefix = '/mnt/lts/sequence_data/rnaseq_data/kn99_database_archive'
coldata_df = read_csv(file.path(archive_prefix, "20220208/combined_df_20220208.csv"))
count_df = read_csv(file.path(archive_prefix, "20220208/counts.csv"))

gene_ids = getGeneNames(database_info$kn99$db_host,
                        database_info$kn99$db_name,
                        Sys.getenv('db_username'),
                        Sys.getenv('db_password'))

gene_ids = gene_ids$gene_id[1:6967]

coldata_df_fltr = coldata_df %>%
  filter(purpose == "fullRNASeq",
         !is.na(fastqFileName))

count_df_fltr = count_df[, colnames(count_df) %in% 
                           coldata_df_fltr$fastqFileName]

coldata_df_fltr = filter(coldata_df_fltr, 
                         fastqFileName %in% 
                           colnames(count_df_fltr))

# make sure the order of the count cols matches order of the metadata rows
count_df_fltr = count_df_fltr[order(match(colnames(count_df_fltr),
                                            coldata_df_fltr$fastqFileName))]

# if this checks false, stop and figure out why. useful functions would be 
# setdiff(). Read the ?setdiff docs -- it is asymetric.
stopifnot(identical(colnames(count_df_fltr), 
                    coldata_df_fltr$fastqFileName))

dds = DESeq2::DESeqDataSetFromMatrix(countData = count_df_fltr,
                                     colData = coldata_df_fltr, 
                                     design = ~1)
rownames(dds) = gene_ids

blrs = brentlabRnaSeqSet(dds = dds)

# everything should work as usual from here
```
```{r}
createEnvPert_epWT = function(blrs){

  # filter
  ep_meta_fastqFileNumbers = extractColData(blrs) %>%
    filter(strain == 'TDY451',
           treatment %in% c("cAMP","noTreatment"),
           treatmentConc %in% c("20", 'noTreatmentConc'),
           experimentDesign == 'Environmental_Perturbation',
           purpose == "fullRNASeq",
           libraryProtocol == 'E7420L',
           !is.na(fastqFileName)) %>%
    pull(fastqFileNumber)

  ep_set = blrs[,colData(blrs)$fastqFileNumber %in%
                  ep_meta_fastqFileNumbers]

  experimental_conditions = alist(medium,
                                  atmosphere,
                                  temperature,
                                  treatment,
                                  treatmentConc,
                                  pH,
                                  timePoint)

  concat_treatment = extractColData(ep_set) %>%
    mutate(concat_treatment = as.factor(paste(!!!experimental_conditions,
                                              sep="_"))) %>%
    pull(concat_treatment)

  colData(ep_set)$concat_treatment = concat_treatment

  ep_set

}
```
