---
title: "Setting Up Your R Environment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SettingUpYouRenvironment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```

# Requirements

An up to date R (so whatever the current version is -- check CRAN) and up to 
date Rstudio

# Install the following

```{r}

cran_pkgs = c('tidyverse', 'devtools')

# note: increase/decrease the number of CPUs. Any machine should be able to 
# handle at least 5. You can probably bump this up to 10 without checking
# anything
install.packages(cran_pkgs, Ncpus = 5)

```

It is possible that there wil be install errors that come up during the 
installation of these two pages. If you are on the computational/analyst side 
of things, then this is a good opportunity to read the error messages and figure 
out what to install -- both packages may have some c or maybe fortran 
dependencies which you will need to install onto your computer. Googling the 
error helps, and frequently R will say, "you need a gcc compiler" or something 
like that.  

If you are not one of the computational/analyst people, then ask for help if 
there are errors in the installation of those two packages. 

# Install the brentlabRnaSeqTools

If there is a 'passing' badge in the github README, then this has been 
successfully built on up to date linux, mac and windows OS. It should also work 
for you. But, if it doesn't, copy the error and make an issue report. Please 
also include the output of `Sys.info()` in your issue report.

```{r}
# as above, you can increase Ncpus
remotes::install.packages("BrentLab/brentlabRnaSeqTools", Ncpus = 5)
```

# Set up some environmental variables

Environmental variables are variables that are read by R and loaded into your 
session when you launch R. You can set either 'user' level variables, which are 
loaded into any R session launched under your current user, and/or you can 
set project level environmental variables, which are only set if you launch into 
a project (see Using Rstudio Projects below). These are convenient to use, and 
are particularly good at avoiding any embarrassing leaks of login credentials 
onto github.  

## User level environmental variables

1. open Rstudio

```{r}
usethis::edit_r_environ('user')
```

2. in the file that is opened, enter some variables that you'd like to have

```{r}
db_username = "some_username"
db_password = "some_password"
```

3. restart your R session

4. see that you can access these variables like so:

```{r}
Sys.getenv("db_username")
# output will be:
#> "some_password"
```

If you put the correct username/password into your environment file, then you 
can access the database like so:

```{r}
library(brentlabRnaSeqTools)

meta = getMetadata(
  database_info$kn99$db_host,
  database_info$kn99$db_name,
  Sys.getenv("db_username"),
  Sys.getenv("db_password"))

```

Since the user level environment file is stored in `~/.Renviron`, there is 
no danger of pushing these credentials up to git if you are in a directory which 
you are tracking with git/github.

# Using an Rstudio project

## set your working directory

If you do not already have a directory in which you store the projects you work 
on, create one. I suggest a direcory called `projects` in your `$HOME`

```{r}
project_dir = "~/projects"
if(!dir.exists(project_dir)){
  dir.create(project_dir)
}
setwd(project_dir)
```

## Create a new Rstudio project

A project is a subset of the infrastructure in an R software package. You can use 
a virtual environment, .Renviron files for environmental variables, etc. It is a 
tool for reproducibility

```{r}
# if this is for the nineyMinuteInduction data, for example, you might call 
# this project "ninetyMinInduction".
project_name = "my_new_project"

usethis::create_project(project_name)
```

## Using the project directory

A new Rstudio session will launch in your new project directory. Now, whenever 
you launch this project, all of your environment variables in .Renviron will be 
loaded. If you are using a virtual environment (a good idea for reproducibility. 
Use `renv`), then your virtual environment will be automatically launched, also.  

The `/R` directory is for R scripts. You could make a `notebooks` directory, or 
just put the notebook in the project parent directory, for example.

[See here for a project directory example](https://github.com/cmatKhan/eds1_mimic_cc)

## Using project level environmental variables in an active R project

Just as with the user level environmental variables, you can set environmental 
variables for a particular project. These are read in addition to the 
user level variables, and will overwrite them if there are two that are named 
the same thing. The first thing to do is to make sure that the project level 
`.Renviron` is in the `.gitignore` so that you don't accidently push up login 
credentials

```{r}
# you can use this, or just click on the .gitignore file in the project
usethis::edit_git_ignore('project')
```

In the `.gitignore` file, add `.Renviron`. Next, open a project level `.Renviron`

```{r}
usethis::edit_r_environ("project")
```

and edit as before. Remember to re-launch your R session after editing the 
`.Renviron` to have access to the environmental variables.

# Using data from the lts archive

```{r}
library(brentlabRnaSeqTools)
library(tidyverse)

# mount the cluster, or download the 20220208 kn99 db archive
# paths will need to be updated for your machine. These are examples
# of what it would look like if you mount.
archive_prefix = '/mnt/lts/sequence_data/rnaseq_data/kn99_database_archive'
coldata_df = read_csv(file.path(archive_prefix, "20220208/combined_df_20220208.csv"))
count_df = read_csv(file.path(archive_prefix, "20220208/counts.csv"))

gene_ids = getGeneNames(database_info$kn99$db_host,
                        database_info$kn99$db_name,
                        Sys.getenv('db_username'),
                        Sys.getenv('db_password'))

gene_ids = gene_ids$gene_id[1:6967]

coldata_df_fltr = coldata_df %>%
  filter(purpose == "fullRNASeq",
         !is.na(fastqFileName))

count_df_fltr = count_df[, colnames(count_df) %in% 
                           coldata_df_fltr$fastqFileName]

coldata_df_fltr = filter(coldata_df_fltr, 
                         fastqFileName %in% 
                           colnames(count_df_fltr))

# make sure the order of the count cols matches order of the metadata rows
count_df_fltr = count_df_fltr[order(match(colnames(count_df_fltr),
                                            coldata_df_fltr$fastqFileName))]

# if this checks false, stop and figure out why. useful functions would be 
# setdiff(). Read the ?setdiff docs -- it is asymetric.
stopifnot(identical(colnames(count_df_fltr), 
                    coldata_df_fltr$fastqFileName))

dds = DESeq2::DESeqDataSetFromMatrix(countData = count_df_fltr,
                                     colData = coldata_df_fltr, 
                                     design = ~1)
rownames(dds) = gene_ids

blrs = brentlabRnaSeqSet(dds = dds)

# everything should work as usual from here
```

