---
title: "Environmental Perturbation Analysis Data Creation"
author: "Chase Mateusiak"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  Create a data object ready for analysis for the Environmental Perturbation 
  experiment set.
  brentlabRnaSeqTools package version: `r packageVersion("brentlabRnaSeqTools")`
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Create an Environmental Perturbation Data Freeze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# BrentlabRnaSeqSet

The `BrentlabRnaSeqSet` is a child of the [DESeqDataSet](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#the-deseqdataset), which is a child of 
[SummarizedExperiment](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html).  

Therefore, the `brentlabRnaSeqSet` has all of the DESeq functionality -- eg, 
DESeq size factor normalization, or the `DESeq()` function -- as well as all of 
the SummarizedExperiment methods, and some more 'custom' methods more suited to 
our purposes. It is easily extensible -- for those more computationally minded 
users, it would be a good idea to learn and use the object oriented programming 
tools. There are many cases in which doing so will make your code less error 
prone, easier to test, more reproducible, and easier to maintain long term.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```

# Setup

```{r setup}
library(brentlabRnaSeqTools)
library(rtracklayer)
library(tidyverse)

# set variables 

KN99_GFF_PATH = Sys.getenv("kn99_gff")
DB_USERNAME = Sys.getenv("db_username")
DB_PASSWORD = Sys.getenv("db_password")

# note: I mount to the cluster and output directly to it
DDS_OUTPUT_DIR = "."

# controls whether dds objects are written
WRITE_OUT = FALSE
```

# Pull the database

```{r}
blrs = brentlabRnaSeqSetFromDatabase('kn99',DB_USERNAME, DB_PASSWORD)

# filter down to just the protein coding genes. Note that this works because the
# nctr-rna annotations were all added to the end of the annotation file.
blrs = blrs[1:6967,]
```

# Add gene level data (optional)

```{r}
# this adds all of the data regarding each locus as a GRange object to the
# gene data slot of the brentlabRnaSeqSet object. Useful if you are going to
# use other Bioconductor packages.

kn99_gff = rtracklayer::import(KN99_GFF_PATH)

kn99_genes = kn99_gff[kn99_gff$ID %in% rownames(blrs)]

rowRanges(blrs) = kn99_genes[order(match(kn99_genes$ID,rownames(blrs)))]

rownames(blrs) = rowData(blrs)$ID
```

# Create Set

If the experiment set you want is not already created, issue an 'issue report', 
and describe the **EXACT** filter that you want to use to create the set. What 
goes into your set depends on what you describe as your filter, so spend time 
with the database to figure out what is there.

```{r}
blrs_ep = createExperimentSet(blrs, 'envPert_epWT')
```

# Quality Filter

```{r}
blrs_ep_qc_passing = qaFilter(blrs_ep, 
                              rle_iqr_threshold = 1, 
                              iqr_colname = "envPert_iqr")
```

# Expression filter

Filter out low expression genes (and anything else you don't want in the 
analysis set). How you do this is up to you, the analyst. [You might use the method 
proposed in this paper](https://www.pnas.org/content/107/21/9546) by, 
among others, Wolfgang Huber, and implemented in the 
[bioconductor package](https://bioconductor.org/packages/release/bioc/html/genefilter.html) 
`geneFilter` (as well as DESeq, by default), or you 
could filter based on some number of samples having greater than some threshold 
of expression, which is shown below.  


```{r}
# How this is done is up to you, and obviously affects what genes are left in.
# Below is an example. You need to think about the thresholds and filter method 
# that suits your purpose best.

expr_fltr <- rowSums(edgeR::cpm(counts(blrs_ep_qc_passing))>4) >= 4

blrs_ep_qc_passing_expr_fltr = blrs_ep_qc_passing[expr_fltr,]
```

# Set Design

```{r}
colData(blrs_ep_qc_passing_expr_fltr)$concat_treatment =
  relevel(colData(blrs_ep_qc_passing_expr_fltr)$concat_treatment,
          ref = 'YPD_noAtmosphere_30_noTreatment_noTreatmentConc_noPH_30')


colData(blrs_ep_qc_passing_expr_fltr) = 
  droplevels(DataFrame(colData(blrs_ep_qc_passing_expr_fltr)))

design(blrs_ep_qc_passing_expr_fltr) = formula(~libraryDate + concat_treatment)
```

# Coerce back to DeseqDataObjects for proessing

```{r}
blrs_ep_qc_passing_expr_fltr = coerceToDds(blrs_ep_qc_passing_expr_fltr)
```


# Write out

```{r}
if(WRITE_OUT){
  today = format(lubridate::today(),"%Y%m%d")

  output_dir = file.path(DDS_OUTPUT_DIR, today)

  dir.create(output_dir, recursive=TRUE)

  write_rds(fltr_full_set_90min$E7420L,
            file.path(output_dir,"envPert_epWT.dds"))
}
```

# Copy the HTCF DESeq scripts to the working directory

These scripts will run DESeq in parallel on HTCF. The only items you'l need to 
edit is the path to the lookup file (if you are running more than one model), 
and to the dds_input in deseq_mpi.sh. If you do not keep deseq_de.R in the same 
directory as deseq_mpi.sh, then you'll need to update the path to the deseq_de.R 
script, also.

```{r}
if(WRITE_OUT){
  htcf_deseq_scripts = c(system.file('bash', 
                            'htcf_parallel_deseq.sh', 
                            package = "brentlabRnaSeqTools"),
                       system.file('R_executable', 
                               'deseq_de.R', 
                                package = "brentlabRnaSeqTools"))

lapply(htcf_deseq_scripts, file.copy, to = DDS_OUTPUT_DIR)
}

```

