---
title: "Using Posit Projects"
author: "Chase Mateusiak"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  A brief introduction to using Posit projects to organize your work.
  brentlabRnaSeqTools package version: `r packageVersion("brentlabRnaSeqTools")`
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Using Posit Projects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```

```{r setup}
library(brentlabRnaSeqTools)
```

# Using an Rstudio project

## set your working directory

If you do not already have a directory in which you store the projects you work 
on, create one. I suggest a directory called `projects` in your `$HOME`

```{r}
project_dir = "~/projects"
if(!dir.exists(project_dir)){
  dir.create(project_dir)
}
setwd(project_dir)
```

## Set your global options

Next, in your Rstudio window, go to Tools -> Global Options. A window will 
open on the `General` tab. I recommend setting your default working directory 
to your new `~/projects` directory. I also strongly recommend that you uncheck 
any box whose description starts with the word "Restore", make sure "Always save 
history" is unchecked, and if there is a dropdown option next to something that 
says "save worksapce to .RData", set that to `never`. You are probably going to 
be moving around a lot of data, and you don't want Rstudio trying to keep track 
of all of it. You're writing code -- if you do it somewhat carefully, it will 
always run and produce the same results. Save the code, don't save intermediate 
data, unless that intermediate data takes a long time to process. Note that you 
should also have an automatic backup set to backup your computer to a remote 
device, and I'd suggest also using github to track changes and as another backup 
for your code.

## Creating a new Posit (Rstudio) project

A project is a subset of the infrastructure in an R software package. You can use 
a virtual environment, .Renviron files for environmental variables, etc. It is a 
tool for reproducibility

```{r}
# if this is for the nineyMinuteInduction data, for example, you might call 
# this project "ninetyMinInduction".
project_name = "my_new_project"

usethis::create_project(project_name)
```

## Using the project directory

A new Rstudio session will launch in your new project directory. Now, whenever 
you launch this project, all of your environment variables in .Renviron will be 
loaded. If you are using a virtual environment (a good idea for reproducibility. 
Use `renv`), then your virtual environment will be automatically launched, also.  

The `/R` directory is for R scripts. You could make a `notebooks` directory, or 
just put the notebook in the project parent directory, for example.

[See here for a project directory example](https://github.com/cmatKhan/eds1_mimic_cc)

## Using project level environmental variables in an active R project

Just as with the user level environmental variables (see the homepage), 
you can set environmental variables for a particular project. 
These are read in addition to the user level variables, and will overwrite 
them if there are two that are named the same thing. The first thing to do is 
to make sure that the project level `.Renviron` is in the `.gitignore` 
so that you don't accidentally push up login credentials to the cloud.

```{r}
# you can use this, or just click on the .gitignore file in the project
usethis::edit_git_ignore('project')
```

In the `.gitignore` file, add `.Renviron`. Next, open a project level `.Renviron`

```{r}
usethis::edit_r_environ("project")
```

and edit as before. Remember to re-launch your R session after editing the 
`.Renviron` to have access to the environmental variables.
