---
title: "QC: Replicate Agreement Object Creation"
author: "Chase Mateusiak"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  Tally the 90minuteInduction experiment set.
  brentlabRnaSeqTools package version: `r packageVersion("brentlabRnaSeqTools")`
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Creating Experiment Set Objects for Replicate Agreement QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```

# Introduction

This illustrates how to construct the data object for the Sample Agreement QC 
step -- RLE -- on each of the data sets. In each case, once you have added the 
design (the last step for each set), you need to write the object to file and 
send it to the cluster. I suggest mounting to the cluster and writing to the 
mounted directory. See the vignette _Running DESeq on HTCF_ for instructions on 
running DESeq on HTCF.  

You can see all defined sets by entering `?createExperimentSet`. If an experiment 
set you are interested in does not exist as an option in `set_names`, then you'll 
need to use the `extractColData(blrs)` of the `blrs` object below to play around 
and come up with a set of filters to define your new set. See the github repository, 
"R/ExperimentSetFunctions.R" for examples

## Setup the environment 

```{r setup}
library(brentlabRnaSeqTools)
library(rtracklayer)
library(tidyverse)
library(caret)

# set variables 

KN99_GFF_PATH = Sys.getenv("kn99_gff")
DB_USERNAME = Sys.getenv("db_username")
DB_PASSWORD = Sys.getenv("db_password")

# note: I mount to the cluster and output directly to it
DDS_OUTPUT_DIR = Sys.getenv("OUTPUT_DIR")
TODAY = format(lubridate::today(),"%Y%m%d")

# controls whether dds objects are written
WRITE_OUT = TRUE
```

## Pull the database

```{r}
blrs = brentlabRnaSeqSetFromDatabase('kn99',DB_USERNAME, DB_PASSWORD)

# filter down to just the protein coding genes. Note that this works because the
# nctr-rna annotations were all added to the end of the annotation file.
blrs = blrs[1:6967,]
```

## Add gene level data (optional)

```{r}
# this adds all of the data regarding each locus as a GRange object to the
# gene data slot of the brentlabRnaSeqSet object. Useful if you are going to
# use other Bioconductor packages, or brentlabRnaSeqTools::createIgvBrowserShot()

kn99_gff = rtracklayer::import(KN99_GFF_PATH)

kn99_genes = kn99_gff[kn99_gff$ID %in% rownames(blrs)]

rowRanges(blrs) = kn99_genes[order(match(kn99_genes$ID,rownames(blrs)))]
```

# 90minuteInduction

## 2016 grant set

This is the data set defined by the single locus KOs in 90minuteInduction 
conditions for genotypes in the `grant_df` (this is a data object which is 
loaded as part of `brentlabRnaSeqTools` -- take a look if you are interested)

### Create the set and quality filter

```{r}
blrs_90min_grant = createExperimentSet(blrs, 'ninetyMin_2016Grant')

# note that this filters out those samples which failed QC1, 
# but but does not filter on RLE unless the argument rle_iqr_threshold 
# is set to a numeric value
blrs_90min_grant_fltr = qaFilter(blrs_90min_grant)

# remove WT which fall on dates with no perturbed samples
blrs_90min_grant_fltr = 
  filterWtByExperimentalLibdate_90min(blrs_90min_grant_fltr)
```

### Add the design

```{r}

blrs_90min_grant_fltr = estimateSizeFactorsByProtocol(blrs_90min_grant_fltr)

min_libdate = min(as.Date(colData(blrs_90min_grant_fltr)$libraryDate))

colData(blrs_90min_grant_fltr)$libraryDate = 
  colData(blrs_90min_grant_fltr)$libraryDate %>%
  relevel(ref = as.character(min_libdate)) %>%
  droplevels()

colData(blrs_90min_grant_fltr)$libraryProtocol = 
  colData(blrs_90min_grant_fltr)$libraryProtocol %>%
  factor() %>%
  relevel(ref = "SolexaPrep") %>%
  droplevels()

colData(blrs_90min_grant_fltr)$genotype1 = 
  colData(blrs_90min_grant_fltr)$genotype1 %>%
  relevel(ref = "CNAG_00000") %>%
  droplevels()

mm = model.matrix(~libraryProtocol+libraryDate+genotype1, 
                  extractColData(blrs_90min_grant_fltr))

min_new_protocol = min(as.Date((unique(pull(filter(
  extractColData(blrs_90min_grant_fltr), 
  libraryProtocol == 'E7420L'), 
  libraryDate)))))

mm = mm[,-which(colnames(mm) == paste0('libraryDate',min_new_protocol))]
```

#### inspect the model matrix for linear dependencies

```{r}
lin_dep_report = caret::findLinearCombos(mm)
```

#### add the design to the object

```{r}
design(blrs_90min_grant_fltr) = mm
rm(mm)

if(WRITE_OUT){
  
  outpath = file.path(DDS_OUTPUT_DIR,
                      "90minDataFreeze",
                      TODAY,
                      "grant_only_input.rds")
  
  dds = DESeq2::DESeqDataSetFromMatrix(
    colData = extractColData(blrs_90min_grant_fltr),
    countData = counts(blrs_90min_grant_fltr),
    design = design(blrs_90min_grant_fltr))
  
  sizeFactors(dds) = sizeFactors(blrs_90min_grant_fltr)
  
  write_rds(dds, outpath)
}
```

## 2016 Grant Set with Doubles

This has all double KO samples in which either of the KO loci are in the `grant_df`, 
plus single KO samples corresponding to one of the double KO perturbed loci and 
WT.

### Create set and quality filter

```{r}
blrs_grant_with_doubles = createExperimentSet(blrs, 'ninetyMin_2016GrantWithDoubles')

# note that this filters out those samples which failed QC1, 
# but but does not filter on RLE unless the argument rle_iqr_threshold 
# is set to a numeric value
blrs_grant_with_doubles_fltr = qaFilter(blrs_grant_with_doubles)

# remove WT which fall on dates with no perturbed samples
blrs_grant_with_doubles_fltr = 
  filterWtByExperimentalLibdate_90min(blrs_grant_with_doubles_fltr)
```

### Add the design

```{r}
blrs_grant_with_doubles_fltr = estimateSizeFactorsByProtocol(blrs_grant_with_doubles_fltr)

min_libdate = min(as.Date(colData(blrs_grant_with_doubles_fltr)$libraryDate))

colData(blrs_grant_with_doubles_fltr)$libraryDate = 
  colData(blrs_grant_with_doubles_fltr)$libraryDate %>%
  relevel(ref = as.character(min_libdate)) %>%
  droplevels()

# add a 'genotype' column which is a concatenation of genotype1 and genotype2
colData(blrs_grant_with_doubles_fltr)$genotype = 
  paste(colData(blrs_grant_with_doubles_fltr)$genotype1,
        colData(blrs_grant_with_doubles_fltr)$genotype2,
        sep = "_") %>%
  str_remove('_$') %>%
  factor() %>%
  relevel(ref = "CNAG_00000") %>%
  droplevels()

colData(blrs_grant_with_doubles_fltr)$libraryProtocol = 
  colData(blrs_grant_with_doubles_fltr)$libraryProtocol %>%
  factor() %>%
  relevel(ref = "SolexaPrep") %>%
  droplevels()

mm = model.matrix(~libraryProtocol+libraryDate+genotype, 
                  extractColData(blrs_grant_with_doubles_fltr))

min_new_protocol = min(
  as.Date(
    (unique(
      pull(
        filter(extractColData(blrs_grant_with_doubles_fltr), 
               libraryProtocol == 'E7420L'), libraryDate)))))

mm = mm[,-which(colnames(mm) == paste0('libraryDate',min_new_protocol))]
```

#### inspect the model matrix for linear dependencies

```{r}
lin_dep_report = caret::findLinearCombos(mm)
lin_dep_report
```

#### add the design to the object

```{r}
design(blrs_grant_with_doubles_fltr) = mm
rm(mm)

if(WRITE_OUT){
  
  outpath = file.path(DDS_OUTPUT_DIR,
                      "90minDataFreeze",
                      TODAY,
                      "grant_doubles_input.rds")
  
  dds = DESeq2::DESeqDataSetFromMatrix(
    colData = extractColData(blrs_grant_with_doubles_fltr),
    countData = counts(blrs_grant_with_doubles_fltr),
    design = design(blrs_grant_with_doubles_fltr))
  
  sizeFactors(dds) = sizeFactors(blrs_grant_with_doubles_fltr)
  
  write_rds(dds, outpath)
}
```

## All

This set is all samples -- singles and doubles -- which are in the 
90minuteInduction conditions. 

### Create set and quality filter

```{r}
blrs_90min_all = createExperimentSet(blrs, 'ninetyMin_all')

# note that this filters out those samples which failed QC1, 
# but but does not filter on RLE unless the argument rle_iqr_threshold 
# is set to a numeric value
blrs_90min_all_fltr = qaFilter(blrs_90min_all)

# remove WT which fall on dates with no perturbed samples
blrs_90min_all_fltr = 
  filterWtByExperimentalLibdate_90min(blrs_90min_all_fltr)
```

### Add the design

```{r}
blrs_90min_all_fltr = estimateSizeFactorsByProtocol(blrs_90min_all_fltr)

min_libdate = min(as.Date(colData(blrs_90min_all_fltr)$libraryDate))

colData(blrs_90min_all_fltr)$libraryDate = 
  colData(blrs_90min_all_fltr)$libraryDate %>%
  relevel(ref = as.character(min_libdate)) %>%
  droplevels()

# add a 'genotype' column which is a concatenation of genotype1 and genotype2
colData(blrs_90min_all_fltr)$genotype = 
  paste(colData(blrs_90min_all_fltr)$genotype1,
        colData(blrs_90min_all_fltr)$genotype2,
        sep = "_") %>%
  str_remove('_$') %>%
  factor() %>%
  relevel(ref = "CNAG_00000") %>%
  droplevels()

colData(blrs_90min_all_fltr)$libraryProtocol = 
  colData(blrs_90min_all_fltr)$libraryProtocol %>%
  factor() %>%
  relevel(ref = "SolexaPrep") %>%
  droplevels()

mm = model.matrix(~libraryProtocol+libraryDate+genotype, 
                  extractColData(blrs_90min_all_fltr))

min_new_protocol = min(
  as.Date(
    (unique(
      pull(
        filter(extractColData(blrs_90min_all_fltr), 
               libraryProtocol == 'E7420L'), libraryDate)))))

mm = mm[,-which(colnames(mm) == paste0('libraryDate',min_new_protocol))]
```

#### inspect the model matrix for linear dependencies

```{r}
lin_dep_report = caret::findLinearCombos(mm)
lin_dep_report
```

#### add the design to the object

```{r}
design(blrs_90min_all_fltr) = mm
rm(mm)

if(WRITE_OUT){
  
  outpath = file.path(DDS_OUTPUT_DIR,
                      "90minDataFreeze",
                      TODAY,
                      "90min_all_input.rds")
  
  dds = DESeq2::DESeqDataSetFromMatrix(
    colData = extractColData(blrs_90min_all_fltr),
    countData = counts(blrs_90min_all_fltr),
    design = design(blrs_90min_all_fltr))
  
  sizeFactors(dds) = sizeFactors(blrs_90min_all_fltr)
  
  write_rds(dds, outpath)
}
```
