---
title: "Rank Reponse Plot"
author: "Chase Mateusiak"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  Create a Rank Response Plot
  brentlabRnaSeqTools package version: `r packageVersion("brentlabRnaSeqTools")`
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Rank Response Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```


```{r setup}
library(brentlabRnaSeqTools)
library(RSQLite)
library(tidyverse)
library(org.Sc.sgd.db)

con = RSQLite::dbConnect(RSQLite::SQLite(),"~/Desktop/cc_metadata/hops_db.sqlite")
```

# Creating a Rank-Response plot

You will need the following:

 - The expression data you wish to use -- there may be more than one (eg, 
 kemmeren data, data from our lab, etc, ...)
 
 - The binding data you wish to use -- again, there may be more than one. For 
 instance, maybe you are comparing chip-chip to calling cards
 
Check `?rank_response_plot` for more information on what columns must exist 
in the expression and binding data.  

An example of creating the plots is below:

```{r}
expr_list = list(
  expr1 = read_tsv("/mnt/scratch/calling_cards/outside_data/yeast_McIsaac_ZEV/YDL106C.DE_15min.txt"),
  expr2 = read_csv("~/Desktop/zev_deseq_res2.csv")
)

name_conversion = read_tsv("/mnt/scratch/calling_cards/outside_data/orf_name_conversion.tab",
                           col_names = c('systematic','common'))

expr1 = read_tsv("/mnt/scratch/calling_cards/outside_data/yeast_McIsaac_ZEV/YOR358W.DE_15min.txt") %>%
  dplyr::rename(systematic = `#gene_sys`, log2FoldChange = shrunken) %>%
  left_join(name_conversion) %>%
  dplyr::select(common, log2FoldChange) %>%
  dplyr::rename(gene = common) %>%
  mutate(padj = 0)

expr1 = read_tsv("/mnt/scratch/calling_cards/outside_data/yeast_Kemmeren_KO/YOR358W.DE.txt") %>%
  dplyr::rename(systematic = `##gene`) %>%
  left_join(name_conversion) %>%
  dplyr::select(common, log2_fold_change,p_value) %>%
  dplyr::rename(gene = common, log2FoldChange = log2_fold_change,padj=p_value)


regions = read_tsv("~/code/callingCardsTools/src/callingcardstools/resources/yeast/orf_coding_all_R61-1-1_20080606.promoter_-700bpto0bp_with_ucsc_seqnames_common_names_coord_corrected.bed", col_names = c('chr', 'start', 'stop', 'gene','score','strand'))
  
binding1 = tbl(con, "regions_yiming_background_adh1_experiment_HAP5_sig") %>%
  filter(sample == "HAP5_run_6021") %>%
  collect() %>%
  left_join(regions) %>%
  dplyr::select(gene,poisson_pval) %>%
  dplyr::rename(binding_signal = poisson_pval) %>%
  left_join(name_conversion, by = c('gene' = 'common')) %>%
  distinct(gene, .keep_all = TRUE)

binding1 = filter(binding1, !gene %in% setdiff(binding1$gene,expr1$gene))
expr1 = filter(expr1, !gene %in% setdiff(binding1$gene,expr1$gene))

binding_list = list(
  binding1 = read_tsv("~/Desktop/t") %>%
    dplyr::rename(gene = `Systematic Name`,
                  binding_signal = `Poisson pvalue`) %>%
    dplyr::select(gene, binding_signal),
  binding2 = read_tsv("~/Desktop/tmp/E0001_PHO4_JP008.sig_prom.txt") %>%
    dplyr::rename(gene = `Systematic Name`,
                  binding_signal = `Poisson pvalue`) %>%
    dplyr::select(gene, binding_signal)
)

rank_response_plot(list(zev = expr1), list(run_6021=binding1), 'test', lfc_thres = 0, padj_thres = .05) +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,.1)) +
  coord_cartesian(xlim = c(0,150))
```

Caveat: I haven't implemented the random response line yet. However, that 
can be added -- what is returned is a ggplot object. Add it with `geom_abline`
