---
title: "UsingRstudioProjects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{UsingRstudioProjects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  include = TRUE,
  eval = FALSE
)
```

# Using an Rstudio project

## install devtools

```{r}
install.packages("devtools")
```

## set your working directory

If you do not already have a directory in which you store the projects you work 
on, create one. I suggest a direcory called `projects` in your `$HOME`

```{r}
project_dir = "~/projects"
if(!dir.exists(project_dir)){
  dir.create(project_dir)
}
setwd(project_dir)
```

## Create a new Rstudio project

A project is a subset of the infrastructure in an R software package. You can use 
a virtual environment, .Renviron files for environmental variables, etc. It is a 
tool for reproducibility

```{r}
# if this is for the nineyMinuteInduction data, for example, you might call 
# this project "ninetyMinInduction".
project_name = "my_new_project"

usethis::create_project(project_name)
```

## Using the project directory

A new Rstudio session will launch in your new project directory. Now, whenever 
you launch this project, all of your environment variables in .Renviron will be 
loaded. If you are using a virtual environment (a good idea. Use `renv`), then 
your virtual environment will be automatically launched, also.  

The `/R` directory is for R scripts. You could make a `notebooks` directory, or 
just put the notebook in the project parent directory, for example.

It is in the `/R` directory that the ninetyMinInduction script should go -- see 
the Quickstart.
